---
title: "Getting started with VariPath"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with VariPath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> Code in this vignette is shown but not executed — the workflow needs a
> Perturb-seq Seurat object that is too large to ship with the package.

## What problem this solves

A Perturb-seq experiment gives you, for every cell, both a perturbation label
and a full transcriptome. The natural question — *which biological pathways did
each perturbation move?* — is awkward to answer from principal components
directly, because PCs are dense: almost every gene loads onto almost every
component, so handing a PC's gene list to an enrichment tool returns something
vague.

VariPath applies a **varimax rotation** to the component space. Varimax
redistributes variance so that each rotated component is loaded heavily by a
smaller set of genes and near-zero by the rest. The subspace spanned is
unchanged; what changes is interpretability. Sparser loadings mean a cleaner
gene set per component, which means a sharper enrichment result — a better
signal-to-noise ratio for perturbation-to-pathway association than the
comparable methods we benchmarked against.

The pipeline is:

```
Seurat object (PCA run)
      │
      ├─ varimax-rotate the PC space           ──►  vari_weights
      ├─ fit perturbation ~ rotated component  ──►  fitted_models, fit_pvals_adj
      └─ enrichR on each component's top genes ──►  adjacency_table
```

## Installation

```{r}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("ssobt/VariPath")
```

Dependencies are declared in `DESCRIPTION`, but a few — `ggtree` and `aplot` in
particular — come from Bioconductor and may need installing by hand:

```{r}
if (!require("BiocManager")) install.packages("BiocManager")
BiocManager::install("ggtree")
install.packages(c("enrichR", "aplot"))
```

A conda environment is recommended if you are managing several Bioconductor
packages at once.

## Input

You need a `Seurat` object that

1. has **PCA already run** (`Seurat::RunPCA()`), and
2. carries a metadata column identifying each cell's perturbation.

```{r}
library(Seurat)
library(VariPath)

# `adata` here is a Seurat object with a 'miR.family' column in its metadata
# giving the decoy / guide identity of each cell.
head(adata@meta.data[["miR.family"]])
```

## Running the analysis

```{r}
out <- sc_varipath(
  seurat_obj          = adata,
  perturbation_column = "miR.family",
  enrichR_dbs         = "GO_Biological_Process_2021",
  loading_cutoff      = 0.1,
  FDR                 = 0.25
)
```

Arguments worth thinking about:

| Argument | Meaning |
|---|---|
| `perturbation_column` | Metadata column holding each cell's perturbation identity. |
| `enrichR_dbs` | Which enrichR libraries to query. Multiple are allowed. |
| `loading_cutoff` | How heavily a gene must load on a rotated component to be included in that component's gene set. Raise it for stricter, smaller sets. |
| `FDR` | Threshold for calling a perturbation-component association significant. |

`sc_varipath()` returns a list:

| Element | Contents |
|---|---|
| `vari_weights` | Varimax gene weights per rotated component. |
| `fitted_models` | Linear models of perturbation against each component. |
| `fit_pvals_adj` | Adjusted p-values for those associations. |
| `perturbation_RC_df` | Tidy perturbation-to-component table, for plotting. |
| `weights_dataframe` | Gene loadings, for plotting. |
| `adjacency_table` | Top three associated pathways per component. |

**Save the result immediately.** The object is expensive to recompute, and the
plotting functions can exhaust memory on large datasets — losing the fit to an
OOM part-way through plotting is an avoidable annoyance:

```{r}
saveRDS(out, "varipath_out.rds")
```

## Plotting

Two views, selected with `plot_type`:

```{r}
# Which perturbations move which rotated components
plot_sc_varipath(out, plot_type = "perturbation")

# Which pathways each rotated component corresponds to
plot_sc_varipath(out, plot_type = "pathway")
```

## Interpreting the output

A rotated component is not a pathway; it is an axis of coordinated expression
that *enriches for* a pathway. Two practical consequences:

- **A perturbation can load onto several components.** That is usually real —
  few perturbations affect exactly one coherent process — rather than a failure
  to converge.
- **Check the gene set, not just the pathway name.** `vari_weights` is there so
  you can see which genes drove an enrichment call before you believe it. A
  component whose top loadings are ribosomal or mitochondrial genes is often
  reporting technical structure rather than biology.

## Related work

`VariPath` came out of the Decoy-seq study, which screened regulatory small
non-coding RNAs by Perturb-seq:

> Choi B\*, Sobti S\*, Soto LM, Charbonneau T, Sababi A, Navickas A,
> Najafabadi HS, Goodarzi H. *Decoy-seq unlocks scalable genetic screening for
> regulatory small noncoding RNAs.* bioRxiv 2025.
> [doi:10.1101/2025.01.25.634869](https://doi.org/10.1101/2025.01.25.634869)

For quantifying *heterogeneity* rather than pathway shifts, see
[txnheterogeneity](https://github.com/ssobt/txnheterogeneity).
